Replication on multicore considering hierarchy and characteristics of the network

### 2013-04-14 paper.tex
Wrote down the algorithm I believe implements atomic broadcast
Fixed the algorithm pseudocode, fixed a label, synced the bibliography
Removed some old stuff

\begin{description}
\item[synchrony] Parts of a multicore machine are synchronous. Hence,
  one characteristic of a node is an identifier for the synchrony
  cluster of a node.
\item[$t_{send}$] Time required to send a message, i.e.\ marshalling
  and queuing the message
\item[$t_{recv}$] Time required to receive a message, i.e.\
  dequeuing, unmarshalling and demultiplexing a channel. 
\end{description}

@@

\begin{algorithm}[H]
  %
  \SetKwInOut{Assumptions}{assumptions}
  \Assumptions{Processes have unique contiguous
    identifiers starting at 0}
  \BlankLine
  %
  \SetKwArray{c}{channels}
  \SetKwFunction{connectNode}{connectNode}
  \SetKwFunction{listen}{listen}
  \SetKwFunction{barrier}{barrier}
  % 
  \KwData{process id $p$, round $r$, %
    each process an array of channels \c}
  \KwResult{Fully-meshed network of processes}
  % 
  \BlankLine
  %
  $r \leftarrow 0$\;
  \For{$i \leftarrow 0$ \KwTo $num(p)$}{
    \eIf{$p=i$}{
      \For{$o \leftarrow i+1$ \KwTo $num(p)$}{
        \c{$o$} $\leftarrow$ \connectNode{$o$}\;
      }
    }{
        \c{$i$} $\leftarrow$ \listen{}\;
    }
    \barrier{} \tcp*{Otherwise, reordering possible}
  }
  \caption{Establish fully-meshed network of channels}
  \label{algo:ab_bindOB}
\end{algorithm}

@@

\begin{algorithm}[H]
\SetKwProg{Fn}{Function}{}{end}%
\SetKwFunction{receive}{receive}%
\SetKwFunction{send}{send}%
  %
  \KwData{List of processes $p$, broadcast tree as graph $(V, E)$}
  \KwResult{Tree based atomic broadcast using a sequentializer}
  % 
  \BlankLine
  \Fn(\tcp*{Receive a message}){\receive{$m$}}{
    $i \leftarrow 5$
    \tcp{Receive a message}
  }
  % 
  \BlankLine
  \Fn(\tcp*{Send a message}){\send{void}}{
    $i \leftarrow 5$
  }
  \caption{Atomic broadcast on reliable communication channels}
  \label{algo:ab}
\end{algorithm}

% How to deduct tree from model?
We then use this graph to build a group-communication
tree. Interesting techniques could be: clustering, to find closely
connected subsets, distributed messages across cluster first, and then
within cluster. Maybe also a spanning-tree protocol, to find a minimal
graph such that all nodes can be reached.

% Tree -> scheduling
Assuming the tree, there is a scheduling problems for the order in
which to send messages to children. The intuition is to send on
long-delay links first to hide the more expensive latencies.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementation}

We implement our quorum protocol on Barrelfish using the UMP
interconnect driver. UMP provides channels between
endpoints. Communication is reliable, and messages are delivered
in-order within a channel. Although UMP channels are uni-directional,
Barrelfish automatically established two uni-directional channels on
bind time. Hence, we consider them as being bi-directional. Every UMP
channels has independent buffers associated with it.

First, we establish connections such that we get a fully connected
graph with vertices being the quorum instances and edges being UMP
channels between them. For that, we need $O(n^2)$ UMP channels.

We build up UMP channels in rounds $r_i$. The quorum instance on core
$i$ is creating a UMP channel to all nodes $j$ with $j>i$ in
$r_i$. The beginning of a round needs to be synchronized such that
node $i+1$ only starts establishing connections when all other nodes
completed $i$. Synchronization is required since UMP channels are
scheduled round-robin. Messages on different channels are received
non-deterministically.

Build a tree. Every node has parents and children (except for the root
and the leafs). The number of children does not really matter for the
algorithm. In the init phase, we construct the tree (e.g. broadcast:
one root node, all other nodes are children of the root node,
multicast: every node has two children, NUMA tree: two layers, two
different implementations?

@@

\paragraph{Measurements to do next} %
\begin{itemize}
\item Play with batching. Incrementally increase the number of
  messages send in a batch. I assume this is machine specific, which
  means, that this is something programmers do not want to deal
  with. We measure this at dynamically at run-time and find the right
  batch size.
\end{itemize}

@@

The average cost is really high (so something is wrong with my
code). But the minimal numbers show what is possible. The numbers
achieved are easily explainable. The cost is increasing
logarithmic with the number of nodes reached by the broadcast (as
expected). Every level in the tree adds an additional 3500 cycles to
the tree. Node 0 takes significantly longer. I don't know yet why that
is.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Open questions}

\paragraph{How does this work with several senders} 
Maybe something like in navigation systems. Every node has an entry
point, which forwards the messages. Nodes are then clustered
\ldots. Or: send the request up to the root and then down again? This
keeps the message complexity the same and doubles the time
complexity.

>>>>

\begin{description}
\item[synchrony] Parts of a multicore machine are synchronous. Hence,
  one characteristic of a node is an identifier for the synchrony
  cluster of a node.
\item[$t_{send}$] Time required to send a message, i.e.\ marshalling
  and queuing the message
\item[$t_{recv}$] Time required to receive a message, i.e.\
  dequeuing, unmarshalling and demultiplexing a channel.
\item[failure] Can nodes fail? If so, how? (Byzantine, \ldots)
\end{description}

@@

\begin{algorithm}[H]
  %
  \SetKwInOut{Assumptions}{assumptions}
  \Assumptions{Processes have unique contiguous
    identifiers starting at 0}
  \BlankLine
  %
  \SetKwArray{c}{channels}
  \SetKwFunction{connectNode}{connectNode}
  \SetKwFunction{listen}{listen}
  \SetKwFunction{barrier}{barrier}
  % 
  \KwData{process id $p$, round $r$, %
    each process an array of channels \c}
  \KwResult{Fully-meshed network of processes}
  % 
  \BlankLine
  %
  $r \leftarrow 0$\;
  \For{$i \leftarrow 0$ \KwTo $num(p)$}{
    \eIf{$p=i$}{
      \For{$o \leftarrow i+1$ \KwTo $num(p)$}{
        \c{$o$} $\leftarrow$ \connectNode{$o$}\;
      }
    }{
        \c{$i$} $\leftarrow$ \listen{}\;
    }
    \barrier{} \tcp*{Otherwise, reordering possible}
  }
  \caption{Establish fully-meshed network of channels}
  \label{algo:ab_bind}
\end{algorithm}

@@

\newcommand{\textc}[1]{{\color{gray} {\footnotesize #1}}}
\begin{algorithm}[H]
\SetCommentSty{textc}
\SetKwInOut{Assumptions}{assumptions}
\Assumptions{Underlying communication channel is reliable and in-order}
\SetKwProg{Fn}{Function}{}{end}%
\SetKwFunction{receive}{on\_receive}%
\SetKwFunction{waitchild}{wait\_for\_children}%
\SetKwFunction{send}{send}%
\SetKwFunction{icsend}{send\_bc\_request}%
\SetKwFunction{icsendack}{send\_bc\_ack}%
\SetKwFunction{handlemessages}{handle\_other\_messages}%
  %
  \KwData{List of processes $p$, broadcast tree as graph $(V, E)$}
  \KwResult{Tree based atomic broadcast using a sequentializer}
  % 
  \BlankLine
  \Fn(\tcp*{Receive a message}){\receive{$client$, $m$}}{
    \For(\tcp*{For all children}){$c \leftarrow \{ c: \exists (self, c) \in E \} $}{
      \icsend{c}
    }
    \waitchild{}\;
    \icsendack{$client$}\;
  }
  % 
  \BlankLine
  \Fn(\tcp*{Send a message}){\send{void}}{
    \tcc{Need to wait for acknowledgment before returning to
      caller. Otherwise, sender might see his own request before some
      other request, that the sequentializer decided to handle first}
    \icsend{$V_{root}$}\tcp*{Relay msg (sequentializer)}
    \While{no answer received}{
      \handlemessages{}\tcp*{Otherwise, deadlocks}
    }
  }
  \caption{Atomic broadcast on reliable communication channels}
  \label{algo:ab}
\end{algorithm}

% How to deduct tree from model?
We then use this graph to build a group-communication
tree. Interesting techniques could be: clustering, to find closely
connected subsets, distributed messages across cluster first, and then
within cluster.

Maybe also a minimum spanning-tree (MST) protocol, to find a minimal
graph such that all nodes can be reached. The problem with the MST is
that it does not consider parallelism. While it is true that the
global cost is minimal in such a tree, it does not take into account
that sending in parallel does not increase the time required to reach
every node. %
\stefan{Find an example, the MST could be a path through the graph
  with weights 1, where it might make more sense to use a link with
  cost 2 in the beginning and then send messages on two ``branches''
  of the tree in parallel.}

% Radix degree
If the send cost $t_s$ is smaller than the propagation time $t_p$,
\stefan{round up}%
$d = \frac{t_p}{t_s}$ messages can be send to hide propagation of
messages. Hence, a tree with degree $d$ should be ideal to minimize
latency, but only for an acknowledged protocol.

% Tree -> scheduling
Assuming the tree, there is a \emph{scheduling problems} for the order in
which to send messages to children. The intuition is to send on
long-delay links first to hide the more expensive latencies, as this
link is dominating total send cost.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example}

@@

@@

The average cost is really high (probably a scheduling issue). But the
minimal numbers show what is possible. The numbers achieved are easily
explainable. The cost is increasing logarithmic with the number of
nodes reached by the broadcast (as expected). Every level in the tree
adds an additional 3500 cycles to the tree. Node 0 takes significantly
longer. I don't know yet why that is.

@@
